### 2010-12-28 Documentation/atropos/atropos.tex
ENH:  A few minor edits.
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@712 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\begin{center}
{\Large \bf An Open Source Framework for Multivariate $n$-Tissue
  Segmentation and Evaluation Using Open Data Sets} \\
\vspace*{0.5cm}
{\normalsize Brian B.~Avants$^{1*}$,  Nicholas J.~Tustison$^2$%
\symbolfootnote[1]{
  The first two authors contributed equally to this work.
}, 
Jue Wu$^1$, Philip A.~Cook$^1$, and James C.~Gee$^1$} \\
\begin{singlespace} 
{\scriptsize  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.\\
  $^2$ Department of Radiology, University of Virginia, Charlottesville, Virginia, USA}
\end{singlespace}
\end{center}

@@

\begin{abstract} 
%Neuroanatomical coordinate systems are essential for the
%interpretation of structural and functional imaging studies.
%However, manual delineation of the sulco-gyral complex is time
%consuming and prone to variability given cortical complexity.  This
%work describes an open source, image-based approach to
%cortical parcellation which uses training data to propagate
%structural labelings to individual images.  Along with other
%neuroinformatics tools publicly distributed in our ANTs (Advanced Normalization Tools)%
We introduce {\em Atropos}, an ITK-based multivariate $n$-class
Expectation Maximization (EM) segmentation tool based on either parametric or non-parametric finite mixture modeling (FMM) of the
class intensities.  Atropos, available in ANTs
\footnote{
http://www.picsl.upenn.edu/ANTs
}, is capable of incorporating spatial prior probability maps,
(sparse) prior label maps and/or Markov Random Field (MRF) modeling.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 69 classes) with a minimal memory footprint.  This work
describes the technical and implementation aspects of Atropos and
evaluates its performance on two different ground-truth datasets.
First, we use the BrainWeb dataset from Montreal Neurological
Institute to evaluate three tissue segmentation performance via (1)
K-means segmentation without use of template data; (2) MRF segmentation with initialization by prior probability maps
derived from a group template; (3) Prior-based segmentation with use
of spatial prior probability maps derived from a group template.  We
also evaluate Atropos performance by using spatial priors to drive a 69 class EM segmentation
problem defined by the Hammers atlas from University College London.  These evaluation studies, combined with
illustrative examples that exercise Atropos options, demonstrate both
performance and wide applicability of this new platform-independent 
open source segmentation tool.
%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.
\end{abstract}

@@

\section{Introduction} As medical image acquisition technology has advanced,
concomitant investment has been made towards adapting classification
techniques for neuroanatomy.  Some of the earliest work appropriated NASA
satellite image processing software for statistical classification of
head tissues in 2-D MR images \citep{Vannier1985}.  A proliferation of
techniques ensued with increasing sophistication in both core
methodology and degree of refinement for specific problems.  The
chronology of progress in segmentation may be tracked through both
technical reviews
\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}
and evaluation studies
\citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.

The problem of accurately delineating the white matter, grey matter
and cerebrospinal fluid of the primate brain continuously spurs
technical development in segmentation.  Following \cite{Vannier1985},
many researchers adopted statistical methods for $n$-tissue anatomical
brain segmentation.  Given the ``missing data'' aspect of this
problem, use of the Expectation-Maximization (EM) framework is natural\citep{Dempster1977}.  The
work described in \cite{Wells1996} was one of the first to use EM for
finding a locally optimal solution by iterating between bias field
estimation and tissue segmentation.  A core component of this work was
explicit modeling of the tissue intensity values as normal
distributions \citep{Cline1990} for both 2-D univariate simulated data
and T1 coronal images, which continues to find utility in contemporary
developments.  A secondary component was an extended nonparametric
probability model, also influenced by earlier work \citep{Kikinis1992}, where
Parzen windowing is used to model the tissue intensity distribution
omitting consideration of the underlying bias field.  Although
technically not an EM-based algorithm, the robustness of the latter has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.

@@

EM may easily be caught in local optima. Consequently, a good
initialization is critical.  Common low-level initialization steps 
include uniform probability assignment \citep{Wells1996},
Otsu thresholding \citep{Zhang2001}, and K-means clustering
\citep{Pappas1992}.  More sophisticated low-level initialization
schemes include that of \cite{Greenspan2006} in which a dense spatial
distribution of Gaussians is used to capture the complex
neuroanatomical layout with subsequent processing used to conjoin
subsets of Gaussians belonging to the same tissue classes.  Recently,
anatomically specific strategies for incorporating prior knowledge
into the segmentation solution have included the use of spatial prior
probability maps of the various structures of interest
\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
probability maps can also be used to provide an initial segmentation.
Related technological developments model partial volume effects for
increased accuracy in brain segmentation
\citep{Ruan2000,Ballester2002,Leemput2003}.

A general trend towards more integrative neuroanatomical image processing led to the work described in
\cite{Ashburner2005} which is publicly available within SPM5, a
large-scale Matlab module in which registration, segmentation, and
bias field correction can be simultaneously modeled within a single
optimization scheme. The roots of this very popular software package
stem back to early work by Karl Friston in which the basis for
statistical parametric mapping was developed \citep{Friston1990} and
subsequently provided to the research community.  Similar integrative brain processing was provided in \cite{Pohl2006} in which
segmentation and registration parameters were optimized simultaneously
while casting the inhomogeneity model parameters of \cite{Wells1996}
as nuisance variables.  Continued work involved recursive parcellation
of the brain volume by considering sub-structures in a hierarchical
manner \citep{Pohl2007}.  An implementation is provided in 3D slicer
\citep{Pieper2006}---an open source package with developmental
contributions from multiple agencies including both private and
academic institutions. Researchers in aging often focus on
accurately segmenting the T1 MRI of elderly controls and subjects
suffering from neurodegeneration, for instance, via SIENA \citep{Smith2007}.  A recent evaluation study 
compared kNN segmentation, SPM Unified Segmentation and SIENA and
found different performance
characteristics under different evaluation criteria
\citep{Bresser2011}.  \cite{Klauschen2009} had similar findings when
comparing SPM5, FSL and FreeSurfer.  These studies suggest that no
single method performs best under every measurement and, along with the No
Free Lunch theorem, highlight the need for segmentation tools that are tunable for different
problems and research goals.

Related neuroanatomical research concerns the selection of geometric
features of the cortex \citep[e.g.][]{Goualher1999} with aims at
understanding the functional-anatomical relationship of the human
brain. Recent endeavors produce a dense cortical labeling in which
every point of the cortex is classified, i.e. a cortical parcellation
\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Given the extreme
time-consumption and tedium associated with manual effort towards the
latter, various techniques have been proposed of which a small subset
has been publicly availed to the research community such as the
popular software package known as Freesurfer
\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
approach detailed in this work, Freesurfer is primarily a
surface-based technique in which the brain structures such as the
grey-white matter interface and pial surfaces are processed, analyzed,
and displayed as tessellated surfaces  \cite{Dale1999,Fischl1999}.
Advantages of surface representations include the ability to map
processed neuroanatomy to simple geometric primitives such as spheres
or planes and the ease of including topological constraints in the
analysis workflow.  These types of methods, as well as Klein's
Mindboggle \cite{Klein2005}, would usually follow an initial segmentation by a volumetric method such as Atropos.

@@

Our open source segmentation tool, which we have dubbed {\em
Atropos}, 
\footnote{ Atropos is one of the three Fates from Greek
mythology characterized by her dreaded shears used to decide the
destiny of each mortal.  Also, consistent with the entomological motif
of our ANTs toolkit, {\it Acherontia atropos} is a species of large
moth known for the skull-like pattern visible on its thorax.  }
efficiently and flexibly implements a $n$-tissue paradigm for
voxel-based image segmentation.  Atropos allows users to harness its
generalized EM algorithm for standard tissue classification of
the brain into gray matter, white matter and cerebrospinal fluid.
Atropos equally allows incarnations that use EM to simultaneously
maximize the posterior probabilities of many classes, for instance,
when parcellating the brain into hemispheres, cortical regions and
deep brain structures such as amygdala, hippocampus and
thalamus.  Atropos contains features of its predecessors for
performing $n$-tissue segmentation including imposition of prior
information in the form of MRFs and template-based spatial prior
probability maps as well as weighted combinations of these terms.  We
also borrow an idea from \cite{Boykov2004} and use sparse spatial
priors to provide initialization and boundary conditions for Atropos
EM segmentation in a semi-interactive manner.  In short, Atropos seeks
to provide a segmentation toolbox that may be modified,
tuned and refined by the user for different problems.

@@

Atropos is multi-platform, free, stand-alone and able to
manage a wide range of scalar and vector image datatypes.  To highlight the
value of this open-source contribution, we performed a search of
software attributes on NITRC and found that as of November 2010 no
stand-alone EM methods are currently listed.
Atropos's memory efficiency enables EM-segmentation of many classes at
once, a capability that is due to a sparse internal representation of
prior probability images that minimizes the need for random access
memory.  We evaluate EM-based classification of $1 mm^3$ T1 MR
neuroimages into 69 neuroanatomical regions to illustrate the practical
value of the low-memory implementation within this paper.  Atropos
also allows multivariate image features to drive the data term, relevant when more than one view of anatomy aids
segmentation, as in neonatal brain tissue classification
\cite{Weisenfeld2009}.  Although Atropos may be applied to
multivariate data from arbitrary modalities, we limit our evaluation
to tissue-classification in T1 neuroimaging in part due to the
abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.  For this reason, all results in this paper are
reproducible with the caveat that users may require some guidance from
the authors or other users in the community.

Organization of this work is as follows: we first describe the theory
behind the various components of Atropos while
acknowledging that theoretical discussion is available elsewhere,
e.g. in previously cited literature.  This is followed by a thorough
discussion of implementation which, though often overlooked, is of
immense practical utility.  We then report evaluation results on
public T1 neuroimaging datasets that have reference gold/silver
standard labelings.  Finally, we provide a discussion of our results
and our open-source contribution in the context of the remainder of
this paper and of previous and future work.

@@

Bayes' theorem provides a powerful mechanism for making inductive
inferences assuming the availability of quantities defining the
relevant conditional probabilities, specifically the likelihood and
prior probability terms.  Bayesian paradigms for brain image
segmentation employ a user-selected observation model defining the
likelihood term and one or more prior probability terms.  The product
of likelihood(s) and prior(s) is proportional to the posterior
probability.  The likelihood term has been previously defined both 
parametrically (e.g. a Gaussian model) and non-parametrically
(e.g. Parzen windowing of the sample histogram).  The prior term, as
given in the literature, has often been formed either as
MRF-based or template-based.  An image segmentation solution in this
context is an assignment of one label to each voxel
\footnote{
In the classic 3-tissue segmentation case, each voxel in the brain region is assigned a label of `cerebrospinal fluid (csf)', `gray matter (gm)', or `white matter (wm)'. 
}
such that the posterior probability is maximized.  The next sections
introduce notation and provide a formal description of three essential
components in Bayesian segmentation, viz.
\begin{itemize}
  \item the likelihood or observation model(s),
  \item the prior probability quantities derived from a generalized
    MRF and template-based prior terms, and 
%, and a novel distance prior used for label propagation, and 
  \item the optimization framework for maximizing the posterior probability.
\end{itemize}
These components are common across most EM segmentation algorithms.

@@

\subsection{Segmentation Objective Function} Atropos optimizes a class
of user-selectable maximization objective functions each of which may
be represented in a generic Bayesian framework, as described by \cite{Sanjay-Gopal1998}.  This framework
requires {\em likelihood models} and {\em prior models} which enter
into Bayes' formula,
\begin{equation}\label{eq:bayes}
p(\mathbf{x}|\mathbf{y})=\underbrace{
p(\mathbf{y}|\mathbf{x})}_{\text{Likelihood(s)}} \underbrace{
p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{\mathbf{y}}, 
\end{equation} 
where the normalization term, $1/\mathbf{y}$, is a constant that does
not affect the optimization \citep{Sanjay-Gopal1998}.
Given choices for likelihood models and prior probabilities, the Bayesian
segmentation solution is the labeling $\hat{\mathbf{x}}$ which
maximizes the posterior probability, i.e.
\begin{align} \hat{\mathbf{x}} = \argmax_{\mathbf{x}}
\left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
\end{align} Similar to its predecessors, Atropos employs
the EM framework \citep{Dempster1977} to find maximum likelihood
solutions to this problem.  The following sections detail the 
Atropos EM along with choices for the likelihood(s), prior(s).

\subsection{Likelihood or Observation Models}
To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.  We denote this set of $K$ likelihood models as $\Phi = \{p_1, p_2, \ldots, p_K\}$.  Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.   Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normal distribution, i.e.
\begin{align}\label{eq:param}
  p_k\left(y_i|l_k\right) &= G\left(\mu_k;\sigma_k\right) \nonumber \\
                    &= \frac{1}{\sqrt{2\pi \sigma_k^2}}\exp\left( \frac{ -(y_i - \mu_k)^2 }{2\sigma_k^2} \right)
\end{align}
where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent
the mean and variance of the $k^{th}$ model.  When $y_i$ is a vector
quantity, we replace the Euclidean distance by by Mahalanobis distance
and define multivariate Gaussian parameters via a mean vector,
$\boldsymbol{\mu}_k$, and covariance matrix, ${\bf \Sigma_k}$.

@@

each iteration as in \cite{Ashburner2005}.

@@

Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.  Due to the large $K$ associated with $K>>1$ problems, exact optimization is even more problematic than for the traditional 3-tissue scenario.

@@

\item[E-Step: ICM Label Update:] Given the initialization and fixed

@@

\item[M-Step: Parameter Update:] 
The posteriors used in this step are called $p^{old}$ to make it
clear that we use the previous parameter estimates.

@@

\gamma^{new}_k \leftarrow  \sum_i \frac{p^{old}_{ik}}{\sum_{j=1}^K p^{old}_{ij}}. 

@@

$$ \mu^{new}_k \leftarrow \frac{ \sum_{i=1}^N  y_i p^{old}_k(l_k|y_i)}{ \sum_{i=1}^N p^{old}_k(l_k|y_i) },
$$
and variance {\bf \textcolor{red}{ FIXME should this be the full
    covariance matrix?}},
$$ \sigma^{new}_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k ) p^{old}_k(l_k|y_i)}{ \sum_{i=1}^N p^{old}_k(l_k|y_i) }.
$$

>>>>

\begin{center}
{\Large \bf An Open Source Framework for Multivariate $n$-Tissue
  Segmentation with Evaluation Using Open Data Sets} \\
\vspace*{0.5cm}
{\normalsize Brian B.~Avants$^{1*}$,  Nicholas J.~Tustison$^2$%
\symbolfootnote[1]{
  The first two authors contributed equally to this work.
}, 
Jue Wu$^1$, Philip A.~Cook$^1$, and James C.~Gee$^1$} \\
\begin{singlespace} 
{\scriptsize  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.\\
  $^2$ Department of Radiology, University of Virginia, Charlottesville, Virginia, USA}
\end{singlespace}
\end{center}

@@

\begin{abstract} 
%Neuroanatomical coordinate systems are essential for the
%interpretation of structural and functional imaging studies.
%However, manual delineation of the sulco-gyral complex is time
%consuming and prone to variability given cortical complexity.  This
%work describes an open source, image-based approach to
%cortical parcellation which uses training data to propagate
%structural labelings to individual images.  Along with other
%neuroinformatics tools publicly distributed in our ANTs (Advanced Normalization Tools)%
We introduce {\em Atropos}, an ITK-based multivariate $n$-class
open source segmentation algorithm distributed
with ANTs.%
\footnote{
http://www.picsl.upenn.edu/ANTs
}  The Bayesian formulation of the segmentation problem is 
solved using the Expectation Maximization (EM) algorithm
with the modeling of the class intensities 
 based on either parametric or non-parametric finite mixtures.  
Atropos is capable of incorporating spatial prior probability maps,
(sparse) prior label maps and/or Markov Random Field (MRF) modeling.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 69 classes) with a minimal memory footprint.  This work
describes the technical and implementation aspects of Atropos and
evaluates its performance on two different ground-truth datasets.
First, we use the BrainWeb dataset from Montreal Neurological
Institute to evaluate three tissue segmentation performance via (1)
K-means segmentation without use of template data; (2) MRF segmentation with initialization by prior probability maps
derived from a group template; (3) Prior-based segmentation with use
of spatial prior probability maps derived from a group template.  We
also evaluate Atropos performance by using spatial priors to drive a 69 class EM segmentation
problem derived from the Hammers atlas from University College London.  These evaluation studies, combined with
illustrative examples that exercise Atropos options, demonstrate both
performance and wide applicability of this new platform-independent 
open source segmentation tool.
%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.
\end{abstract}

@@

\section{Introduction} As medical image acquisition technology has advanced,
significant investment has been made towards adapting classification
techniques for neuroanatomy.  Early work appropriated NASA
satellite image processing software for statistical classification of
head tissues in 2-D MR images \citep{Vannier1985}.  A proliferation of
techniques ensued with increasing sophistication in both core
methodology and degree of refinement for specific problems.  The
chronology of progress in segmentation may be tracked through both
technical reviews
\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}
and evaluation studies
\citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.

The problem of accurately delineating the white matter, grey matter
and cerebrospinal fluid (and further subdivisions) of the human brain continuously spurs
technical development in segmentation.  Following \cite{Vannier1985},
many researchers adopted statistical methods for $n$-tissue anatomical
brain segmentation.  Given the ``missing data'' aspect of this
problem, use of the Expectation-Maximization (EM) framework is natural \citep{Dempster1977}.  The
work described in \cite{Wells1996} was one of the first to use EM for
finding a locally optimal solution by iterating between bias field
estimation and tissue segmentation.  A core component of this work was
explicit modeling of the tissue intensity values as normal
distributions \citep{Cline1990} for both 2-D univariate simulated data
and T1 coronal images, which continues to find utility in contemporary
developments.  A secondary component was an extended nonparametric
probability model, also influenced by earlier work \citep{Kikinis1992}, where
Parzen windowing is used to model the tissue intensity distribution
omitting consideration of the underlying bias field.  Although
technically not an EM-based algorithm, the robustness of the latter has motivated its continued use even more recently \citep[e.g.][]{Weisenfeld2009}.

@@

Several initialization strategies have been proposed to overcome the characteristic
susceptibility of EM algorithms to local optima.   Common low-level initialization steps 
include uniform probability assignment \citep{Wells1996},
Otsu thresholding \citep{Zhang2001}, and K-means clustering
\citep{Pappas1992}.  More sophisticated low-level initialization
schemes include that of \cite{Greenspan2006} in which a dense spatial
distribution of Gaussians is used to capture the complex
neuroanatomical layout with subsequent processing used to conjoin
subsets of such Gaussians belonging to the same tissue classes.  Recently,
anatomically specific strategies for incorporating prior knowledge
into the segmentation solution have included the use of spatial prior
probability maps of the various structures of interest
\citep{Leemput1999a,Marroquin2002,Ashburner2005}.  These spatial prior
probability maps can also be used to provide an initial segmentation.
Related technological developments model partial volume effects for
increased accuracy in brain segmentation
\citep{Ruan2000,Ballester2002,Leemput2003}.

A general trend towards more integrative neuroanatomical image processing led to the work described in
\cite{Ashburner2005} which is publicly available within SPM5, a
large-scale Matlab module in which registration, segmentation, and
bias field correction can be simultaneously modeled within a single
optimization scheme. The roots of this very popular software package
stem back to early work by Karl Friston which laid the basis for
statistical parametric mapping \citep{Friston1990}. 
 Similar integrative brain processing was provided in \cite{Pohl2006} in which
segmentation and registration parameters were optimized simultaneously
while casting the inhomogeneity model parameters of \cite{Wells1996}
as nuisance variables.  Continued work involved recursive parcellation
of the brain volume by considering sub-structures in a hierarchical
manner \citep{Pohl2007}.  An implementation is provided in 3D slicer
\citep{Pieper2006}---an open source medical image computation and
visualization package with developmental
contributions from multiple agencies including both private and
academic institutions.

Related neuroanatomical research concerns the selection of geometric
features of the cortex \citep[e.g.][]{Goualher1999} with aims at
understanding the functional-anatomical relationship of the human
brain. Recent endeavors produce a dense cortical labeling in which
every point of the cortex is classified, i.e. a cortical parcellation
\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Given the extreme
time-consumption and tedium associated with manual effort towards the
latter, various techniques have been proposed of which a small subset
has been publicly availed to the research community such as the
popular software package known as Freesurfer
\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
approach detailed in this work, Freesurfer is primarily a
surface-based technique in which the brain structures such as the
grey-white matter interface and pial surfaces are processed, analyzed,
and displayed as tessellated surfaces  \citep{Dale1999,Fischl1999}.
Advantages of surface representations include the ability to map
processed neuroanatomy to simple geometric primitives such as spheres
or planes and the ease of including topological constraints in the
analysis workflow.  These types of methods, including Klein's
Mindboggle \citep{Klein2005}, would usually follow an initial segmentation by a 
volumetric method such as Atropos.

@@

Researchers in aging often focus on
accurately segmenting the T1 MRI of elderly controls and subjects
suffering from neurodegeneration, for instance, via SIENA \citep{Smith2007}.  A recent evaluation study 
compared kNN segmentation, SPM Unified Segmentation and SIENA and
found different performance
characteristics under different evaluation criteria
\citep{Bresser2011}.  \cite{Klauschen2009} had similar findings when
comparing SPM5, FSL and FreeSurfer.  These studies suggest that no
single method performs best under every measurement and, along with the No
Free Lunch theorem \citep{Wolpert1997}, highlight the need for segmentation tools that are tunable for different
problems and research goals.

Our open source segmentation tool, which we have dubbed {\em
Atropos},%
\footnote{Atropos is one of the three Fates from Greek
mythology characterized by her dreaded shears used to decide the
destiny of each mortal.  Also, consistent with the entomological motif
of our ANTs toolkit, {\it Acherontia atropos} is a species of large
moth known for the skull-like pattern visible on its thorax.  }
efficiently and flexibly implements a $n$-tissue paradigm for
voxel-based image segmentation.  Atropos allows users to harness its
generalized EM algorithm for standard tissue classification of
the brain into gray matter, white matter and cerebrospinal fluid even
in cases of multivariate image data---relevant when more than one 
view of anatomy aids segmentation, as in neonatal brain tissue classification
\citep[e.g.][]{Prastawa2005,Weisenfeld2009}.  
Atropos equally allows incarnations that use EM to simultaneously
maximize the posterior probabilities of many classes with minimal
random access memory requirements, for instance,
when parcellating the brain into hemispheres, cortical regions and
deep brain structures such as amygdala, hippocampus and
thalamus.  Atropos contains features of its predecessors for
performing $n$-tissue segmentation including imposition of prior
information in the form of MRFs and template-based spatial prior
probability maps as well as weighted combinations of these terms.  We
also borrow an idea from \cite{Boykov2004} and use sparse spatial
priors to provide initialization and boundary conditions for Atropos
EM segmentation in a semi-interactive manner.  In short, Atropos seeks
to provide a segmentation toolbox that may be modified,
tuned and refined for different use scenarios.

@@

%Atropos is multi-platform, free, /stand-alone and able to
%manage a wide range of scalar and vector image datatypes.

To highlight the
value of this open-source contribution, we performed a search of
software attributes on NITRC and found that as of November 2010 no
stand-alone EM methods are currently listed.
%Atropos's memory efficiency enables EM-segmentation of many classes at
%once, a capability that is due to a sparse internal representation of
%prior probability images that minimizes the need for random access
%memory.  
We evaluate EM-based classification of $1 mm^3$ T1 MR
neuroimages into 69 neuroanatomical regions to illustrate the practical
value of the low-memory implementation within this paper.  
Although Atropos may be applied to
multivariate data from arbitrary modalities, we limit our evaluation
to tissue-classification in T1 neuroimaging in part due to the
abundance of ``gold-standard'' data for this modality.  Consistent
with our advocacy of open science (not to mention the facilitation of
analysis due to accessibility) we also only use publicly available
data sets.  For this reason, all results in this paper are
reproducible with the caveat that users may require some guidance from
the authors or other users in the community.

Organization of this work is as follows: we first describe the theory
behind the various components of Atropos while
acknowledging that additional theoretical discussion is available elsewhere.  
This is followed by a thorough
discussion of implementation which, though often overlooked, is of
immense practical utility.  We then report evaluation results on
public T1 neuroimaging datasets that have reference gold/silver
standard labelings.  Finally, we provide a discussion of our results
and our open source contribution in the context of the remainder of
this paper and of previous and future work.

@@

Bayes' theorem provides a powerful mechanism for making inductive
inferences assuming the availability of quantities defining the
relevant conditional probabilities, specifically the likelihood and
prior probability terms.  Bayesian paradigms for brain image
segmentation employ a user-selected observation model defining the
likelihood term and one or more prior probability terms.  The product
of likelihood(s) and prior(s) is proportional to the posterior
probability.  The likelihood term has been previously defined both 
parametrically (e.g. a Gaussian model) and non-parametrically
(e.g. Parzen windowing of the sample histogram).  The prior term, as
given in the literature, has often been formed either as
MRF-based and/or template-based.  An image segmentation solution in this
context is an assignment of one label to each voxel%
\footnote{
In the classic 3-tissue segmentation case, each voxel in the brain region is assigned a label of `cerebrospinal fluid (csf)', `gray matter (gm)', or `white matter (wm)'. 
}
such that the posterior probability is maximized.  The next sections
introduce notation and provide a formal description of three essential
components in Bayesian segmentation, viz.
\begin{itemize}
  \item the likelihood or observation model(s),
  \item the prior probability quantities derived from a generalized
    MRF and template-based prior terms, and 
%, and a novel distance prior used for label propagation, and 
  \item the optimization framework for maximizing the posterior probability.
\end{itemize}
These components are common across most EM segmentation algorithms.

@@

\subsection{Segmentation Objective Function}

 Atropos optimizes a class
of user-selectable objective functions each of which may
be represented in a generic Bayesian framework, as described by \cite{Sanjay-Gopal1998}.  This framework
requires {\em likelihood models} and {\em prior models} which enter
into Bayes' formula,
\begin{equation}\label{eq:bayes}
p(\mathbf{x}|\mathbf{y})=\underbrace{
p(\mathbf{y}|\mathbf{x})}_{\text{Likelihood(s)}} \underbrace{
p(\mathbf{x})}_{\text{Prior(s)}}\frac{1}{\mathbf{y}}, 
\end{equation} 
where the normalization term, $1/\mathbf{y}$, is a constant that does
not affect the optimization \citep{Sanjay-Gopal1998}.
Given choices for likelihood models and prior probabilities, the Bayesian
segmentation solution is the labeling $\hat{\mathbf{x}}$ which
maximizes the posterior probability, i.e.
\begin{align} \hat{\mathbf{x}} = \argmax_{\mathbf{x}}
\left\{p(\mathbf{y}|\mathbf{x})p(\mathbf{x})\right\}.
\end{align} Similar to its predecessors, Atropos employs
the EM framework \citep{Dempster1977} to find maximum likelihood
solutions to this problem.  The following sections detail the 
Atropos EM along with choices for the likelihood and prior terms.

\subsection{Likelihood or Observation Models}
To each of the $K$ labels corresponds a single probabilistic model describing the variation of $\mathcal{F}$ over $\mathcal{I}$.  We denote this set of $K$ likelihood models as $\Phi = \{p_1, p_2, \ldots, p_K\}$.  Using the standard notation, $\mathrm{Pr}(S=s) = p(s)$, $\mathrm{Pr}(S=s|T=t) = p(s|t)$, we can define these voxelwise probabilities, $\mathrm{Pr}_k( Y_i = y_i | X_i = l_k ) = p_k(y_i|l_k)$, in either parametric or non-parametric terms.   Given its simplicity and good performance, in the parametric case, $p_k$ is typically defined as a normal distribution, i.e.
\begin{align}\label{eq:param}
  p_k\left(y_i|l_k\right) &= G\left(\mu_k;\sigma_k\right) \nonumber \\
                    &= \frac{1}{\sqrt{2\pi \sigma_k^2}}\exp\left( \frac{ -(y_i - \mu_k)^2 }{2\sigma_k^2} \right)
\end{align}
where the parameters $\mu_k$ and $\sigma_k^2$ respectively represent
the mean and variance of the $k^{th}$ model.  When $y_i$ is a vector
quantity, we replace the Euclidean distance by by Mahalanobis distance
and define multivariate Gaussian parameters via a mean vector,
$\boldsymbol{\mu}_k$, and covariance matrix, ${\bf \Sigma}_k$.

@@

each iteration as in \cite{Ashburner2005}.%

@@

Consider $N$ sites each with a possible $K$ labels for a total of $N^K$ possible labeling configurations.  Due to the large $K$ associated with $K>>1$ problems, exact optimization is even more intractable than for the traditional 3-tissue scenario.

@@

\item[ICM Label Update (E-Step):] Given the initialization and fixed

@@

\item[Parameter Update (M-Step):] 
Note that the posteriors used in the previous iteration are used to 
estimate the parameters at the current iteration.  

@@

\gamma_k \leftarrow \frac{1}{N} \sum_{i=1}^N  p_k(l_k|y_i)
%\gamma^{new}_k \leftarrow  \sum_i \frac{p^{old}_{ik}}{\sum_{j=1}^K p^{old}_{ij}}. 

@@

 \mu_k \leftarrow \frac{ \sum_{i=1}^N  y_i p_k(l_k|y_i)}{ \sum_{i=1}^N p_k(l_k|y_i) },
and variance,
 \sigma^2_k \leftarrow  \frac{ \sum_{i=1}^N  (y_i - \mu_k ) p_k(l_k|y_i)}{ \sum_{i=1}^N p_k(l_k|y_i) }.
The latter two quantities are modified, respectively, in the case of multivariate data as follows:
 \boldsymbol{\mu}_k \leftarrow \frac{ \sum_{i=1}^N  \mathbf{y}_i p_k(l_k|\mathbf{y}_i)}{ \sum_{i=1}^N p_k(l_k|\mathbf{y}_i) },
and the $(m,n)^{th}$ element of the covariance matrix, $\boldsymbol{\Sigma}$, is calculated from
 \boldsymbol{\Sigma}_{mn,k} \leftarrow \frac{ \sum_{i=1}^N  ( \mathbf{y}_i - \boldsymbol{\mu}_k ) p_k(l_k|\mathbf{y}_i)}{1 - \sum_{i=1}^N p^2_k(l_k|\mathbf{y}_i) }.
